---
title: "Introduction to dimensionality reduction"
output:
  html_document:
    df_print: paged
---

### Overview

This notebook deals with a linear (plane) and non-linear (sphere) 2D manifolds embedded in R^3 feature space. It applies PCA, kernel PCA and other non-linear methods to reduce the input data to 2D transformed space. It compares the outcomes of the individual dimensionality reduction methods. The comparison is in terms of their reconstruction error, if the reconstruction mapping is available or purely visually if the mapping is not available.

### Load libraries, define the reconstruction error

The reconstruction error is the sum of Euclidean distances between the original samples and their reconstructions in the input space. The sample reconstructions are reached through reconstruction mapping from the (reduced) transformed space, certain information (noise) may be lost.

```{r prepare, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(scatterplot3d) # 3d plots
library(mvtnorm) # generate spheres
library(kernlab) # kernel PCA

#' Calculate mean reconstruction error
#'
#' @param x the original sample matrix (samples in rows, features in columns)
#' @param r the sample matrix after reconstruction (samples in rows, features in columns), the same dimension as x
#' @return the mean reconstruction error
#' @export
recError <- function(x, r){
  if (sum(dim(x)==dim(r))!=2){stop("Different dimensions!")}
  recError<-0
  for (i in 1:nrow(x)){
    recError<-recError+sqrt(sum((x[i,]-r[i,])^2))
  }
  return(recError/nrow(x))
}
```

### Generate input data 

Generates two manifolds. The first is linear, it is a plane embedded in R^3 space (3D). The second is non-linear, it is a sphere, again embedded in R^3 space. 

```{r}
D = 3         # number of input dimensions
m = 200       # number of samples
sigma = 0.2   # standard deviation for noisy manifolds

x<-runif(m,-1,1)
y<-runif(m,-1,1)
plane<-cbind(x,y,z=-x-0.5*y) # the plane is: x+0.5y+z=0, the normal vector is (1,0.5,1) or (0.66,0.33,0.66) when having unit size
plane<-plane[order(plane[,"x"]+plane[,"y"]),] # sort observations to be able to visually understand them
noisyPlane<-plane
noisyPlane[,"z"]<-noisyPlane[,"z"]+rnorm(n=m,mean=0,sd=sigma)

scatterplot3d(plane,color=rainbow(nrow(plane))) 
s3d<-scatterplot3d(noisyPlane,color=rainbow(nrow(plane)))
recoveredPlane<-lm(z~.,as.data.frame(noisyPlane)) # relearn the plane from noisy data, should approach the plane
s3d$plane3d(recoveredPlane, lty="solid")  # add the recovered plane to the plot with noisy plane
noisy.3d <- s3d$xyz.convert(noisyPlane[,"x"],noisyPlane[,"y"],noisyPlane[,"z"])  # locations of data points
flat.3d <- s3d$xyz.convert(noisyPlane[,"x"],noisyPlane[,"y"], fitted(recoveredPlane)) # corresponding locations on the regression plane
segments(noisy.3d$x, noisy.3d$y, flat.3d$x, flat.3d$y, lty="dashed")  # draw lines from data points to plane

dirs <- rmvnorm(n = m, mean = rep(0, D)) # sample
colnames(dirs) <- c("x","y","z")
sphere <- dirs / sqrt(rowSums(dirs^2)) # normalize
sphere <- sphere[order(sphere[,"z"]+cos(sphere[,"x"])+cos(sphere[,"y"])),] # sort observations to be able to visually understand them
noisySphere <- sphere + rmvnorm(n = m, mean = rep(0, D), sigma = sigma/20 * diag(D)) # add noise

scatterplot3d(sphere,color=rainbow(m))
scatterplot3d(noisySphere,color=rainbow(m))
```
## Run PCA in first case

Run PCA for the first manifold, reduce the dimension to the intrinsic dimension equal to 2, reconstruct the data and check its reconstruction error. The script shows that PCA captures the manifold, dimensionality reduction maintains a major part of variance and may partly denoise the samples.
```{r}
pca_noisyPlane<-prcomp(noisyPlane) # the most common call
eigen(cov(noisyPlane)) # an analogy of the previous call, calculate eigenvectors from the covariance matrix
summary(pca_noisyPlane) # get info about projection, the first two components capture a major part of variance!
biplot(pca_noisyPlane,cex=0.7) # understand the data, the role of features in principal components and distribution of objects
scatterplot3d(pca_noisyPlane$x,color=rainbow(nrow(plane)),zlim = c(-2,2),type = "h") # Full transformation
plot(pca_noisyPlane$x[,1:2],col=rainbow(nrow(plane))) # once more with the sample colors, PC1 and PC2 maintain the plane (shape, positions)

# see the projection in the tranformed space, no reduction
t_noisyPlane<-noisyPlane%*%pca_noisyPlane$rotation
t_noisyPlane_centered<-as.data.frame(t(apply(t_noisyPlane,1,"-",colMeans(t_noisyPlane)))) # center the data
head(predict(pca_noisyPlane,noisyPlane)) # the same result with predict from prcomp

# now reduce the data and see what happens in the input space
t_noisyPlane_reduced<-cbind(t_noisyPlane[,c(1,2)],rep(0,m))

r_noisyPlane<-t_noisyPlane%*%t(pca_noisyPlane$rotation) # identical with the original values, to check and show the reconstruction
r_noisyPlane_reduced<-t_noisyPlane_reduced[,c(1,2)]%*%t(pca_noisyPlane$rotation[,c(1,2)]) # get the reconstruction after 2D reduction
r_noisyPlane_reduced<-as.matrix(noisyPlane)%*%pca_noisyPlane$rotation[,c(1:2)]%*%t(pca_noisyPlane$rotation[,c(1,2)]) # the same outcome, a different formula

s3d<-scatterplot3d(noisyPlane)
s3d$points3d(r_noisyPlane_reduced,col=rainbow(m),pch=4)
s3d$plane3d(recoveredPlane, lty="solid")  # add the recovered plane to the plot with noisy plane
recovered.3d <- s3d$xyz.convert(r_noisyPlane_reduced[,"x"],r_noisyPlane_reduced[,"y"],r_noisyPlane_reduced[,"z"])  # locations of data points
flat.3d <- s3d$xyz.convert(r_noisyPlane_reduced[,"x"],r_noisyPlane_reduced[,"y"], fitted(recoveredPlane)) # corresponding locations on the regression plane
segments(recovered.3d$x, recovered.3d$y, flat.3d$x, flat.3d$y, lty="dashed")  # draw lines from data points to plane
```

When looking at the reconstruction error, let us see the error caused by noise in the input space. This error serves as a reference. Its value is `r recError(plane,noisyPlane)`. Then, we may easily check that reconstruction with no dimensionality reduction fully recovers the original noisyPlane, the reconstruction error is `r recError(noisyPlane,r_noisyPlane)`. Next, we can see that noisyPlane is still well captured after the reduction, the reconstruction error is `r recError(noisyPlane,r_noisyPlane_reduced)`. Eventually, let us compare the original plane wihout any noise and the plane recovered after the dimensionality reduction. This error is `r recError(plane,r_noisyPlane_reduced)`. If smaller than the error in the noisyPlane (the first one), dimensionality reduction works as (partial) denoising.

There is a couple of reasons why denoising is not perfect in terms of our reconstruction error. Firstly, the plane is learnt from a limited sample, it is not recovered perfectly. Compare the normal vector of the original plane (0.66,0.33,0.66) and the normal vector of the denoised plane given by PC3 (the perpendicular vector to the recovered plane captured by PC1 and PC2) which is `r pca_noisyPlane$rotation[,"PC3"]`. Secondly, denosing means to map to any point that lies on the manifold, that ncessarily does not have to be the original point (compare the two error definitions in the lecture).

## Run PCA for the second manifold

The procedure remains the same. The script shows that PCA captures the manifold, dimensionality reduction maintains a major part of variance and may partly denoise the samples.
```{r}
pca_noisySphere<-prcomp(noisySphere) # the most common call
eigen(cov(noisySphere)) # an analogy of the previous call, calculate eigenvectors from the covariance matrix
summary(pca_noisySphere) # get info about projection, the first two components capture only slightly more than 2/3 of variance!
biplot(pca_noisySphere,cex=0.7) # understand the data, the role of features in principal components and distribution of objects

# see the projection in the tranformed space, no reduction
t_noisySphere<-noisySphere%*%pca_noisySphere$rotation
t_noisySphere_centered<-as.data.frame(t(apply(t_noisySphere,1,"-",colMeans(t_noisySphere)))) # center the data
head(predict(pca_noisySphere,noisySphere)) # the same result with predict from prcomp
scatterplot3d(t_noisySphere_centered,color=rainbow(m))

# now reduce the data and see what happens in the input space
t_noisySphere_reduced<-cbind(t_noisySphere[,c(1,2)],rep(0,m))

r_noisySphere<-t_noisySphere%*%t(pca_noisySphere$rotation) # identical with the original values, to check and show the reconstruction
r_noisySphere_reduced<-t_noisySphere_reduced[,c(1,2)]%*%t(pca_noisySphere$rotation[,c(1,2)]) # get the reconstruction after 2D reduction
r_noisySphere_reduced<-as.matrix(noisySphere)%*%pca_noisySphere$rotation[,c(1:2)]%*%t(pca_noisySphere$rotation[,c(1,2)]) # the same outcome, a different formula

s3d<-scatterplot3d(sphere)
s3d$points3d(r_noisySphere_reduced,,col=rainbow(m),pch=4)
recovered.3d <- s3d$xyz.convert(r_noisySphere_reduced[,"x"],r_noisySphere_reduced[,"y"],r_noisySphere_reduced[,"z"])  # locations of data points
orig.3d <- s3d$xyz.convert(sphere[,"x"],sphere[,"y"], sphere[,"z"]) # the original sphere
segments(recovered.3d$x, recovered.3d$y, orig.3d$x, orig.3d$y, lty="dashed")  # draw lines from data points to Sphere
```

TYe referential error caused purely by the noise is in this spherical case `r recError(sphere,noisySphere)`. Similarly to the previous example, the reconstruction with no dimensionality reduction fully recovers the original noisySphere, the reconstruction error is `r recError(noisySphere,r_noisySphere)`. On the contrary to the previous example, the dimensionality reduction leads to the large reconstruction errors. The noisySphere reduction error is much larger than the original noise, its magnitude is `r recError(noisySphere,r_noisySphere_reduced)`. At the same time, no denoising happened as the error that captures the distance between the original sphere without noise and the sphere recovered after the dimensionality reduction is `r recError(sphere,r_noisySphere_reduced)`.

## Apply kernel PCA

The noise removal in the second case can obviously be improved with non-linear methods such as kernel PCA. We will try kernel PCA with three different kernels ... linear (no change from PCA), polynomial and RBF.

```{r}
kpc <- kpca(~.,data=as.data.frame(noisySphere),kernel="polydot",kpar=list(degree=1,offset=0),features=3) # run kernel PCA with linear kernel
scatterplot3d(pcv(kpc),color=rainbow(m),main="Linear kernel") # identical with PCA, the manifold not denoised
kpc <- kpca(~.,data=as.data.frame(noisySphere),kernel="polydot",kpar=list(degree=2),features=3) # run kernel PCA with quadratic kernel
scatterplot3d(pcv(kpc),color=rainbow(m),main="Quadratic kernel") # improper transform
kpc <- kpca(~.,data=as.data.frame(noisySphere),kernel="rbfdot",kpar=list(sigma=0.8),features=3) # run kernel PCA with RBF kernel
# print the principal component vectors
scatterplot3d(pcv(kpc),color=rainbow(m),main="RBF kernel") # the manifold is denoised, however the object positions changed
plot(pcv(kpc)[,1:2],col=rainbow(m)) # the first two components only 
```

