---
title: "Statistical data analysis"
subtitle: "Spectral Clustering"
author: "Dominik Hoftych"
date: "November 05, 2018"
output: html_document
---

```{r}
rm(list=ls())
setwd("~/Dropbox/CVUT/MGR/1. semestr/SAN/5_cv_san")
```

```{r include=FALSE}
LoadDataset <- function(path, title="Unknown dataset", csv = FALSE){
  if(csv){
    X <- read.csv(path, header = FALSE)
  }
  else{
    X <- read.table(path) 
  }
  names(X) <- c("x", "y", "label")
  plot(X$x, X$y, col=X$label, xlab = "x", ylab = "y", main = title)
  return(X)
}

CalcSimMatrix <- function(points, sigma = 0.5){
  D <- dist(points, method = "euclidean") #Euclidean distance matrix
  D <- as.matrix(D)
  D <- D^2
  S <- exp(-D / (2 * sigma^2)) #Gaussian similarity function
  return(S)
}

PlotConnectedGraph <- function(data, dist){
  plot(data$x, data$y, col=data$label, xlab = "x", ylab = "y", main = "Connectivity graph")
  indecesx <- which(dist > 0, arr.ind = TRUE)
  for(i in 1:nrow(indecesx)){
    lines(data[indecesx[i,],])
  }
}

#Get accuracy
Purity <- function(clusters, labels){
  #clusters - vector of values determining the class
  #labels
  return(sum(apply(table(labels, clusters), 2, max)) / length(labels))
}

```

#Load a data
In this task, we will work with two synthetic datasets: Spiral dataset (spiral.txt) containing three clusters and Jain's toy problem dataset (jain.txt) containing two clusters. Now, we will demonstrate a code for the first dataset (Spiral dataset). A code and results for the second dataset have to be attached by students at the end of this report. Firstly, look at the original Spiral dataset with the ideal assignment.
```{r load data}
#Dataset
#mydataset <- LoadDataset("jain.txt", "A.K. Jain's Toy problem")
mydataset <- LoadDataset("spiral.txt", "Spiral dataset")
```

Visualization of k-means algorithm that was applied to the Spiral dataset. Purity function shows how good each point was assigned to the proper cluster.
```{r kmeans}
#Kmeans
res <- kmeans(mydataset, centers = 3)
plot(mydataset$x, mydataset$y, col=res$cluster, main = "Spiral dataset - k-means", xlab = "x", ylab = "y")
Purity(res$cluster, mydataset$label)
```

#1.Computation of similarity matrix
CalcSimMatrix function calculates Euclidean distances between pairs of points and afterwards applies Gaussian kernel.
```{r}
#Calculate Similarity matrix
d <- CalcSimMatrix(mydataset[,1:2], sigma = 0.35) # 0.15
```

#2.Construction of similarity graph
In this task, we take into consideration three different similarity graphs. All of them, you will implement.

###2.a $\epsilon$-graph [0.5p]
The first and the most trivial is $\epsilon$-graph that connects all points whose pairwise distance are smaller than $\epsilon$.
```{r}
BuildEpsilonGraph <- function(D, e = 0.2){
  D[D<e] <- 0
  return(D)
}
```

For checking and showing a connectivity graph use 'PlotConnectedGraph' function.
```{r}
#Epsilon Graph
similarityGraph_epsilon <- BuildEpsilonGraph(d, e = 0.002)
PlotConnectedGraph(mydataset, similarityGraph_epsilon)

```

###2.b Directed knn graph [1p]
Directed graph is a such graph that connects vertex $v_i$ with vertex $v_j$ if $v_j$ is among the $k$-nearest neighbors of $v_i$. Note that this leads to non-symetric relationships.
```{r}
BuildDirectedKNNGraph <- function(D, k = 5){
  # one way
  # indicesCols<- as.matrix(t(apply(D,2,order,decreasing = TRUE)))
  # firstKindsCols <- indicesCols[,1:k]
  # for (i in 1:dim(D)[1]) {
  #    D[i,-c(firstKindsCols[i,])] <- 0
  # }
  
  # another way
  sortedRowInds<- as.matrix(t(apply(D,1,order,decreasing = TRUE)))
  firstKcolumnsInds <- sortedRowInds[,1:k]
   for (i in 1:dim(D)[1]) {
     for(j in 1:dim(D)[2]){
        if(!is.element(j,firstKcolumnsInds[i,])){
          D[i,j] <- 0
        }
      }
   }
  return(D)
}

```

```{r}
#Directed Graph
similarityGraph_direct <- BuildDirectedKNNGraph(d, k = 3)
# print(similarityGraph_direct[1:5,1:5])
PlotConnectedGraph(mydataset, similarityGraph_direct)

```

###2.c Undirected knn graph [1p]
To making similarity graph undirected, we can apply two approaches. First variant connect $v_i$ and $v_j$ if $v_i$ is among the $k$-nearest neighbors of $v_j$ or if $v_j$ is among the $k$-nearest neighbors of $v_i$. The second variant connect vertices $v_i$ and $v_j$ if both $v_i$ is among the $k$-nearest neighbors of $v_j$ and $v_j$ is among the $k$-nearest neighbors of $v_i$. Use 'mutual' parameter in 'BuildUndirectedKNNGraph' function to distinguish between these two variants.
```{r}
BuildUndirectedKNNGraph <- function(D, k = 5, mutual = FALSE){
  
  sortedRowInds<- as.matrix(t(apply(D,1,order,decreasing = TRUE)))
  firstKcolumnsInds <- sortedRowInds[,1:k]
  
  # not used
  # sortedColInds <- as.matrix(t(apply(D,2,order,decreasing=TRUE)))
  # firstKrowsInds <- sortedColInds[1:k,]
  
  for(i in 1:dim(D)[1]){
    if(mutual){
      for(j in 1:dim(D)[2]){
        # I really dont know why this is not working.....
        # if(!is.element(j,firstKcolumnsInds[i,]) | !is.element(i,firstKrowsInds[,j])){
        #   D[i,j] <- 0
        # }
        if(!is.element(j,firstKcolumnsInds[i,]) | !is.element(i,firstKcolumnsInds[j,])){
           D[i,j] <- 0
        }
      }
    }else{
      for(j in 1:dim(D)[2]){
        # I really dont know why this is not working.....
        # if(!is.element(j,firstKcolumnsInds[i,]) & !is.element(i,firstKrowsInds[,j])){
        #   D[i,j] <- 0
        # }
        if(!is.element(j,firstKcolumnsInds[i,]) & !is.element(i,firstKcolumnsInds[j,])){
           D[i,j] <- 0
        }
      }
    }
  }
  return(D)
}
```

The first variant of undirected knn graph.
```{r}
similarityGraph_undirect_false <- BuildUndirectedKNNGraph(d, k = 3, mutual = FALSE)
print(similarityGraph_undirect_false[1:5,1:5])
# print(similarityGraph_undirect_false[100:125,100:125])
PlotConnectedGraph(mydataset, similarityGraph_undirect_false)

```

The second variant of undirected knn graph.
```{r}

similarityGraph_undirect_true <- BuildUndirectedKNNGraph(d, k = 5, mutual = TRUE)
print(similarityGraph_undirect_true[1:5,1:5])
# print(similarityGraph_undirect_true[100:125,100:125])
PlotConnectedGraph(mydataset, similarityGraph_undirect_true)
```

#3. Derivation of Laplace matrix [0.5p]
Now, it is necessary to implement Laplace matrices of the similarity graph.
```{r}
CalcLaplacian <- function(W){
  # calc degree matrix D - diagonal matrix containing degrees of each vertex
  degrees <- rowSums(W)
  D <- diag(degrees)
  # calc laplacian matrix
  L = D - W
  return(L)
}
```


#4. Plotting Spiral dataset

After Laplace matrix was computed, we compute eigenvectors and select the last (smallest) $k$ eigenvectors. Finally, we apply k-means algorithm on the last $k$ eigenvectors and show final plot using 'plot' function and also we check the purity score.
```{r}
#laplacian matrix
laplacian_epsilon <- CalcLaplacian(similarityGraph_epsilon)
laplacian_direct <- CalcLaplacian(similarityGraph_direct)
laplacian_undirect_false <- CalcLaplacian(similarityGraph_undirect_false)
laplacian_undirect_true <- CalcLaplacian(similarityGraph_undirect_true)
#Compute eigenvectors
eigenVectors_epsilon <- eigen(laplacian_epsilon)
eigenVectors_direct <- eigen(laplacian_direct)
eigenVectors_undirect_false <- eigen(laplacian_undirect_false)
eigenVectors_undirect_true <- eigen(laplacian_undirect_true)
#Get last (smallest) k eigenvectors
finalEigenvector_epsilon <- eigenVectors_epsilon$vectors[,c(length(eigenVectors_epsilon$values), length(eigenVectors_epsilon$values)-1)]
finalEigenvector_direct <- eigenVectors_direct$vectors[,c(length(eigenVectors_direct$values), length(eigenVectors_direct$values)-1)]
finalEigenvector_undirect_false <- eigenVectors_undirect_false$vectors[,c(length(eigenVectors_undirect_false$values), length(eigenVectors_undirect_false$values)-1)]
finalEigenvector_undirect_true <- eigenVectors_undirect_true$vectors[,c(length(eigenVectors_undirect_true$values), length(eigenVectors_undirect_true$values)-1)]

#Kmeans
resSpectral_epsilon <- kmeans(centers = 3, finalEigenvector_epsilon, nstart = 20, iter.max = 100)
resSpectral_direct <- kmeans(centers = 3, finalEigenvector_direct, nstart = 20, iter.max = 100)
resSpectral_undirect_false <- kmeans(centers = 3, finalEigenvector_undirect_false, nstart = 20, iter.max = 100)
resSpectral_undirect_true <- kmeans(centers = 3, finalEigenvector_undirect_true, nstart = 20, iter.max = 100)
#Plot data
plot(mydataset$x, mydataset$y, col=resSpectral_epsilon$cluster, main = "Spiral dataset - Spectral Clustering, epsilon", xlab = "x", ylab = "y")
Purity(resSpectral_epsilon$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_direct$cluster, main = "Spiral dataset - Spectral Clustering, direct", xlab = "x", ylab = "y")
Purity(resSpectral_direct$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_undirect_false$cluster, main = "Spiral dataset - Spectral Clustering, undirect, false", xlab = "x", ylab = "y")
Purity(resSpectral_undirect_false$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_undirect_true$cluster, main = "Spiral dataset - Spectral Clustering, undirect, true", xlab = "x", ylab = "y")
Purity(resSpectral_undirect_true$cluster, mydataset$label)
```

#5. Experimental results [1p]
Summarize your experience from Spiral dataset experiment (what option is important, what option does not have influence, etc.).

> : For epsilon graph, I had to lower the value of epsilon in order to achieve purity 1
  : For both directed and unmutual undirected graph, I had to lower the value of k.. 
  : In case of mutual undirectred graph it was not necessary and purity 1 was achieved with k=5 as well
  : Since purity was achieved just with lowering the value of k, I did not even try to play with kmeans iterations

Do the same procedure for Jain's Toy problem dataset, i.e. plot all graphs with interesting results. Also, summarize your experience from experiments (what option is impor-
tant, what option does not have influence, etc.).

```{r findTheBestk, eval=FALSE}
#Dataset
mydataset <- LoadDataset("jain.txt", "A.K. Jain's Toy problem")
#Calculate Similarity matrix
d <- CalcSimMatrix(mydataset[,1:2], sigma = 0.35) # 0.15
```

```{r}
BuildEpsilonGraph <- function(D, e = 0.2){
  D[D<e] <- 0
  return(D)
}
```

```{r}
#Epsilon Graph
similarityGraph_epsilon <- BuildEpsilonGraph(d, e = 0.002)
PlotConnectedGraph(mydataset, similarityGraph_epsilon)
```

```{r}
BuildDirectedKNNGraph <- function(D, k = 5){
  sortedRowInds<- as.matrix(t(apply(D,1,order,decreasing = TRUE)))
  firstKcolumnsInds <- sortedRowInds[,1:k]
   for (i in 1:dim(D)[1]) {
     for(j in 1:dim(D)[2]){
        if(!is.element(j,firstKcolumnsInds[i,])){
          D[i,j] <- 0
        }
      }
   }
  return(D)
}

```

```{r}
#Directed Graph
similarityGraph_direct <- BuildDirectedKNNGraph(d, k = 7)
PlotConnectedGraph(mydataset, similarityGraph_direct)
```

```{r}
BuildUndirectedKNNGraph <- function(D, k = 5, mutual = FALSE){
  sortedRowInds<- as.matrix(t(apply(D,1,order,decreasing = TRUE)))
  firstKcolumnsInds <- sortedRowInds[,1:k]

  for(i in 1:dim(D)[1]){
    if(mutual){
      for(j in 1:dim(D)[2]){
        if(!is.element(j,firstKcolumnsInds[i,]) | !is.element(i,firstKcolumnsInds[j,])){
           D[i,j] <- 0
        }
      }
    }else{
      for(j in 1:dim(D)[2]){
        if(!is.element(j,firstKcolumnsInds[i,]) & !is.element(i,firstKcolumnsInds[j,])){
           D[i,j] <- 0
        }
      }
    }
  }
  return(D)
}
```

```{r}
similarityGraph_undirect_false <- BuildUndirectedKNNGraph(d, k = 10, mutual = FALSE)
PlotConnectedGraph(mydataset, similarityGraph_undirect_false)

similarityGraph_undirect_true <- BuildUndirectedKNNGraph(d, k = 8, mutual = TRUE)
PlotConnectedGraph(mydataset, similarityGraph_undirect_true)
```

```{r}
#laplacian matrix
laplacian_epsilon <- CalcLaplacian(similarityGraph_epsilon)
laplacian_direct <- CalcLaplacian(similarityGraph_direct)
laplacian_undirect_false <- CalcLaplacian(similarityGraph_undirect_false)
laplacian_undirect_true <- CalcLaplacian(similarityGraph_undirect_true)
#Compute eigenvectors
eigenVectors_epsilon <- eigen(laplacian_epsilon)
eigenVectors_direct <- eigen(laplacian_direct)
eigenVectors_undirect_false <- eigen(laplacian_undirect_false)
eigenVectors_undirect_true <- eigen(laplacian_undirect_true)
#Get last (smallest) k eigenvectors
finalEigenvector_epsilon <- eigenVectors_epsilon$vectors[,c(length(eigenVectors_epsilon$values), length(eigenVectors_epsilon$values)-1)]
finalEigenvector_direct <- eigenVectors_direct$vectors[,c(length(eigenVectors_direct$values), length(eigenVectors_direct$values)-1)]
finalEigenvector_undirect_false <- eigenVectors_undirect_false$vectors[,c(length(eigenVectors_undirect_false$values), length(eigenVectors_undirect_false$values)-1)]
finalEigenvector_undirect_true <- eigenVectors_undirect_true$vectors[,c(length(eigenVectors_undirect_true$values), length(eigenVectors_undirect_true$values)-1)]

#Kmeans
resSpectral_epsilon <- kmeans(centers = 2, finalEigenvector_epsilon, nstart = 20, iter.max = 100)
resSpectral_direct <- kmeans(centers = 2, finalEigenvector_direct, nstart = 20, iter.max = 100)
resSpectral_undirect_false <- kmeans(centers = 3, finalEigenvector_undirect_false, nstart = 20, iter.max = 100)
resSpectral_undirect_true <- kmeans(centers = 3, finalEigenvector_undirect_true, nstart = 20, iter.max = 100)
#Plot data
plot(mydataset$x, mydataset$y, col=resSpectral_epsilon$cluster, main = "A.K. Jain's Toy problem dataset - Spectral Clustering, epsilon", xlab = "x", ylab = "y")
Purity(resSpectral_epsilon$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_direct$cluster, main = "A.K. Jain's Toy problem dataset - Spectral Clustering, direct", xlab = "x", ylab = "y")
Purity(resSpectral_direct$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_undirect_false$cluster, main = "A.K. Jain's Toy problem dataset - Spectral Clustering, undirect, false", xlab = "x", ylab = "y")
Purity(resSpectral_undirect_false$cluster, mydataset$label)
plot(mydataset$x, mydataset$y, col=resSpectral_undirect_true$cluster, main = "A.K. Jain's Toy problem dataset - Spectral Clustering, undirect, true", xlab = "x", ylab = "y")
Purity(resSpectral_undirect_true$cluster, mydataset$label)
```

> : In case of epsilon graph, I did not even manage to achieve purity 1 - both lowering the eps value and sigma didn't work (maybe I tried wrong combinations of them)
  : For directed graph, increasing the value of k helped
  : For undirected graphs, I had to increase both number of clusters and the value of k, to achieve purity 1
  : Generally, this datased was harder to optimize and needed either higher k values or higher number of clusters


