---
title: "MidtermTest"
author: "Matous Dzivjak"
output: pdf_document
---
  
# Arrests

```{r}
advertising <-  read.csv('datasets/Advertising.csv', header = TRUE)
arrests <-  read.csv('datasets/USArrests.csv', header = TRUE)
advertising <- as.data.frame(advertising)
summary(arrests)
attach(arrests)
quantile(Murder, probs = c(0.05), names=TRUE)
boxplot(Rape, main = "Boxplot of rape", xlab = "All states", ylab = "Number of rapes")
```

In boxplot the box has center on median (half of the scores are greater and half lower) and 
lower and upper bounds where 50% of the scores lay (e.g. 50% of the scores are in the box). The lines connected to the box with vertical lines
are upper quartile (seventy-five percent of the scores fall below the upper quartile) and 
lower quartile (twenty-five percent of scores fall below the lower quartile). 
Separate points are outliers 

# Advertising

```{r}
library(ggplot2)
attach(advertising)
summary(advertising)
```

## TV sales
```{r}
TVmodel <- lm(sales ~ TV)
summary(TVmodel)
plot(sales, TV, col = "blue", main = "Dependence of Sales on TV advertisment", xlab = "TV Advertisment budget", ylab = "Sales")
ggplot(data = advertising) + 
  geom_point(mapping = aes(x = TV, y = sales)) +
  geom_smooth(mapping = aes(x = TV, y = sales), method=lm) 
```

## Newspaper sales
```{r}
newsmodel <- lm(sales ~ newspaper)
summary(newsmodel)
plot(sales, newspaper, col = "blue", main = "Dependence of Sales on News advertisment", xlab = "Newspaper Advertisment budget", ylab = "Sales")
ggplot(data = advertising) + 
  geom_point(mapping = aes(x = newspaper, y = sales)) +
  geom_smooth(mapping = aes(x = newspaper, y = sales), method=lm) 
```

## Predict sales
```{r}
newdata <- data.frame(TV = c(50))
predict(TVmodel, newdata)

newdata <- data.frame(newspaper = c(50))
predict(newsmodel, newdata)
```

We expect greater increase for newspaper because for one unit of advertisment we have bigger change in sales as cen be seen in the coefficient from summary of the model.