---
title: "Confirmation analysis"
author: "Dominik Hoftych"
date: "November 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())

knitr::opts_chunk$set(echo = TRUE)

#load the dataset
Load <- function(path = "potato.dat"){
  potatoData <- read.table(path, header = FALSE)
  colnames(potatoData) <- c("Area", "Size" ,"Temp", "Period", "Method", "Texture", "Flavor", "Moistness")
  potatoData$Area <- as.factor(potatoData$Area)
  potatoData$Size <- as.factor(potatoData$Size)
  potatoData$Temp <- as.factor(potatoData$Temp)
  potatoData$Period <- as.factor(potatoData$Period)
  potatoData$Method <- as.factor(potatoData$Method)
  return(potatoData)
}

source("boxTest.R")
library(mvnormtest)
library(MVN)

#load the dataset
potatoData <- Load()
potatoData_method1 <- potatoData[which(potatoData[,5]==1),]
potatoData_method2 <- potatoData[which(potatoData[,5]==2),]
potatoData_method3 <- potatoData[which(potatoData[,5]==3),]
potatoData_method4 <- potatoData[which(potatoData[,5]==4),]
potatoData_method5 <- potatoData[which(potatoData[,5]==5),]
```


# 1. Null hypothesis

H\_o = Quality of potatoes is not influenced by its cooking method.
Level of statistical significance used for testing is \alpha = 5% = 0.05.



# 2. Visualize data
```{r}

## PCA 
pc <- prcomp(potatoData[,c(6:8)], center = TRUE,scale = TRUE)
summary(pc)
pca12 <- pc$x[,1:2] 
pca12 <- cbind(pca12, potatoData[,5])
colnames(pca12)[3] <- "Cooking method"

#plot
plot(pca12[,1], pca12[,2], col = pca12[,3],
     main = "Potato dataset in 2d", xlab = "PCA1", ylab = "PCA2")

library(ggplot2)
# convert matrix to data frame
pcadf <- as.data.frame(pca12)
pcadf$`Cooking method` <- as.factor(pcadf$`Cooking method`)

ggplot(pcadf, aes(x=PC1,y=PC2,color=`Cooking method`)) + ggtitle("Potato dataset in 2d") + geom_point() + scale_color_manual(values = c('red','blue', 'green','black', 'purple')) 

# biplot
biplot(pc)
```

# 3. Check assumptions
```{r}
# 1. independence - granted
# 2. random sampling granted 
# 3. Multivariate normality
mshapiro.test(t(potatoData_method1[,c(6,7,8)]))
mshapiro.test(t(potatoData_method2[,c(6,7,8)]))
mshapiro.test(t(potatoData_method3[,c(6,7,8)]))
mshapiro.test(t(potatoData_method4[,c(6,7,8)]))
mshapiro.test(t(potatoData_method5[,c(6,7,8)]))

# 4. Homogeneity of covariance matrices
source("boxTest.R")
BoxMTest(potatoData[,c(6,7,8)], cl = potatoData$Method)
```


# 4. Testing my hypothesis
```{r}
# MANOVA
potato_manova <- manova(as.matrix(potatoData[,6:8]) ~ Method, data = potatoData)
summary(potato_manova, intercept = TRUE, test = "Pillai")
summary(potato_manova, intercept = TRUE, test = "Wilks")
summary(potato_manova, intercept = TRUE, test = "Hotelling")
summary(potato_manova, intercept = TRUE, test= "Roy")

# ANOVA
potato_anova_Texture <- anova(lm(as.matrix(potatoData[,6]) ~ Method, data = potatoData))
potato_anova_Flavor <- anova(lm(as.matrix(potatoData[,7]) ~ Method, data = potatoData))
potato_anova_Moistness <- anova(lm(as.matrix(potatoData[,8]) ~ Method, data = potatoData))

# another way, result is the same
aov_Texture <- aov(as.matrix(potatoData[,6]) ~ Method, data=potatoData)
anova(aov_Texture)
aov_Flavor <- aov(as.matrix(potatoData[,7]) ~ Method, data=potatoData)
anova(aov_Flavor)
aov_Moistness <- aov(as.matrix(potatoData[,8]) ~ Method, data=potatoData)
anova(aov_Moistness)
```


# 5. Summary

According to observed value of the tested statistics with level of statistical significance \alpha = 5% = 0.05, we deny the null hypothesis in behalf of the alternative hypothesis. The conclusion is that the cooking method influences at least one criterium that is used to evaluate the quality of potatoes.









